\documentclass[12pt]{article}
\def\baselinestretch{1.0}
\setlength{\topmargin}{-40pt}
\setlength{\textheight}{660pt}
\setlength{\oddsidemargin}{-40pt}
\setlength{\evensidemargin}{20pt}
\setlength{\textwidth}{527pt}
\setlength{\footskip}{30pt}
\parindent 0.3in
\hyphenpenalty=10000
\tolerance=10000
\pagestyle{empty}

\begin{document}
\noindent
Genome 562 \hfill  Population Genetics \\
Winter, 2017 \hfill  J. Felsenstein

\centerline{\bf Homework 1}
\centerline{\bf due January 13, 2017}

\bigskip

Do these problems, and email the results (if you can, in a PDF or else
in Word .DOC or .DOCX format).  Like all of the weekly homeworks this one
totals 25 points and is due at the end of that day (at midnight).

\bigskip

\begin{enumerate}
\item (12 points)   In a population we take a sample of 200 individuals and
find these numbers of the genotypes at a 3-allele locus:
\begin{center}
\begin{tabular}{c c c c c c}
$A_1 A_1$ & $A_1 A_2$ & $A_1 A_3$ & $A_2 A_2$ & $A_2 A_3$ & $A_3 A_3$ \\
\hline
10 & 16 & 84 & 6 & 52 & 32 \\
\end{tabular}
\end{center}
\begin{description}
\item[(i)] What are the gene frequencies of the alleles in the sample?
\item[(ii)] What genotype frequencies do we expect for the genotypes under
random mating at those gene frequencies?
\item[(iii)] Is there an excess of heterozygotes in all heterozygous
genotypes?  Is there an overall excess of heterozygotes?  Is there an excess of
homozygotes for all homozygous genotypes?
\item[(iv)] To what extent could these genotype frequencies be explained by
having different gene frequencies in the female and the male parents? {\it
Hint: you can do this by just messing about until you figure it out, rather than by algebra.}
\end{description}
\bigskip

\bigskip

\item (13 points)  Suppose that at two loci, a large population has two haplotypes,
$AB$ and $ab$ (there are initially no $Ab$ or $aB$ haplotypes). The two loci
are unlinked, so they segregate independently in any double-heterozygote.
The initial haplotype frequencies are 0.6 and 0.4 in both sexes.
If it is a random-mating population,
\begin{description}
\item[(i)] What are the gene frequencies of the $A$ and the $B$ alleles?
\item[(ii)] What haplotype frequencies do we predict if there is no
linkage disequilibrium?
\item[(iii)] What is the value of the linkage disequilibrium parameter $D$ ?
\item[(iv)] What will its value be a generation from now?  Will it be zero?
Why or why not?
\end{description}

\end{enumerate}

\end{document}

