\documentclass[11pt]{article}
\def\baselinestretch{0.7}
\setlength{\topmargin}{-60pt}
\setlength{\textheight}{700pt}
\setlength{\oddsidemargin}{0pt}
\setlength{\evensidemargin}{60pt}
\setlength{\textwidth}{477pt}
\setlength{\footskip}{0pt}
\parindent 0.3in
\hyphenpenalty=10000
\tolerance=10000
\pagestyle{empty}

\begin{document}

{\parindent=0in
Genome 562 \hfill  Population Genetics \\
Winter, 2017 \hfill  J. Felsenstein
\vfill

\centerline{\bf \Large Supplementary questions}
\bigskip

(These are to be answered in the numbered questions on the back of the main
mark-sense form, {\it not on
the yellow form}.  Use the letters on a scale of A = agree strongly, through G = disagree
strongly.)
\bigskip

\begin{enumerate}
{\Large
\item The text was useful.
\item The text was good.
\item The course was too concentrated on theory.  There should have been much
more discussion of data.
\item The course should have exams instead of homeworks.
\item The computer simulation program was helpful.
\item The homeworks should have been more challenging.
\item The availability of the audio recordings of lectures was helpful.
\item Lectures based on projections would have been better for me than having
Joe write on the board in ``chalk-talk'' format.
\item The course covered the material too quickly.
\item I really don't need to know this material.
\item I already knew most of this material.
\item The course should have concentrated on how to use programs that analyze
population-genetic data instead of on the theory.
}
\end{enumerate}
}
\vfill

\vfill


\end{document}

